---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Preparing the environment

```{r set_options}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')

library(data.table)
library(ggplot2)
```


## Loading and preprocessing the data

```{r read_data}
getwd()
mydata <- read.csv("activity.csv", header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
```

```{r process}
mydata$date <- as.Date(mydata$date, format = "%Y-%m-%d")
mydata$interval <- as.factor(mydata$interval)
```


## What is mean total number of steps taken per day?

```{r pre_calc_stepsperday}
steps_taken <- aggregate(steps ~ date, mydata, sum)
colnames(steps_taken) <- c("date","steps")
head(steps_taken)
```

The Histogram of the total number of steps taken per day:

```{r histo}
ggplot(steps_taken, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps taken per day", 
             x = "Number of Steps per Day",
             y = "Number of times in a day (Count)")  
```

Mean of the number of steps taken per day:

```{r meanmedian}
steps_mean   <- round(mean(steps_taken$steps, na.rm=TRUE), digits=2)
steps_median <- round(median(steps_taken$steps, na.rm=TRUE), digits=2)
steps_mean
steps_median
```


## What is the average daily activity pattern?


```{r daily_activity}

daily_activity <- aggregate(mydata$steps, 
                            by = list(interval = mydata$interval),
                            FUN=mean, na.rm=TRUE)
daily_activity$interval <- 
        as.integer(levels(daily_activity$interval)[daily_activity$interval])
colnames(daily_activity) <- c("interval", "steps")
 
ggplot(daily_activity, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average Daily Activity Pattern",
             x="Interval",
             y="Number of steps") 

max_interval <- daily_activity[which.max(daily_activity$steps),]
max_interval
```


## Imputing missing values

```{r missing_values}
missing_vals <- sum(is.na(mydata$steps))
```

The total number of ***missing values*** are **`r missing_vals`**.

```{r missing_value_imputation}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

mydata_fill <- data.frame(  
        steps = na_fill(mydata, daily_activity),  
        date = mydata$date,  
        interval = mydata$interval)

sum(is.na(mydata_fill$steps))

fill_steps_per_day <- aggregate(steps ~ date, mydata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", 
             y = "Number of times in a day(Count)") 


steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
steps_mean_fill
steps_median_fill
```


## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
    
    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)
    
    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
    
    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(mydata_fill)
```


```{r plot_weekdays}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="red") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps")
        

```

